### 2015-09-07 paper/chemokine_main_revised.tex
clean-up to JTB inital state

\documentclass[10pt]{article}

@@

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

@@

% Use the PLoS provided bibtex style
\bibliographystyle{plos2009}

@@

\usepackage{color}
\usepackage[usenames,dvipsnames,table]{xcolor}
\usepackage{ulem}

@@

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Chemokines versus viral replication: a spatial of model of lymphocyte search in the influenza infected lung}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$,
Candice Clay$^{2}$, 
Judy Cannon$^{3}$,
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
\bf{3} Department of Molecular Genetics \& Microbiology Department of Pathology, University of New Mexico Health Sciences Center, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.  Viral clearance depends in part on the ability of T cells to locate the infection in time to prevent uncontrolled spread of the virus.  We use an agent-based spatial model to quantify how efficiently T cells locate and control different strains of influenza in the lung.  We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature.  The spatial model reveals challenges for T cell recruitment not apparent in standard differential equation models.  The model shows that virus kinetics largely determine the course of the infection, while T cell chemokine parameters have less effect.

@@

% ---- 300 word abstract

% Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.  Subtle differences in how a given strain interacts with the immune system are likely a key factor in determining virulence.  One aspect of the interaction is the ability of T cells to locate the foci of the infection in time to prevent uncontrolled expansion.  Here, we focus on T cell navigating from the lymph nodes through the vascular system to the infection, using an agent-based spatial model to investigate how different strains of influenza influence the efficiency of this process.

%We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature when available.  The spatial nature of the model reveals unique challenges for T cell recruitment that are not apparent in standard differential equation models. As plaques expand in size, the ratio of infected cells to dead cells decreases, making it difficult for T cells to encounter infected cells efficiently.  Further, differences in diffusion rates between chemokines and virions can misdirect T cells to areas that no longer contain infected epithelial cells.  Sensitivity analysis of the model identifies the viral parameters as the most significant factor in determining the course of the resulting infection.  Finally, we use the model parameters to determine an upper bound on T cell search time in the model.

% ---- end 300 word abstract

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
% \section*{Author Summary}

@@

Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \cite{Who2009}.  Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the the mechanisms that control this variability are not well understood.  Pathogenicity is a function of the virus and its interaction with the host immune response.

Identifying the critical components of this complex process is key to understanding viral pathogenicity and designing therapeutic interventions.  However, determining the relative contribution of these myriad experimental components is challenging. We address this challenge with computational modeling, which gives us the advantage of studying relative contributions of different aspects of the host immune response and accounts for strain-specific viral dynamics in the model.

Rather than developing a comprehensive immune system model we focus on one aspect of the adaptive response specifically, studying how activated antigen-specific CD8 T cells migrate from the lymph node to the focus of infection (FOI) in the human lung.  We assume that chemotactic gradients play an important role in guiding T cells to the FOI and that T cells resist extravasation and recirculate in the absence of inflammatory signals.  Chemotactic proteins are known to enhance T cell recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005}, where recruited T cells mediate either protection or immunopathology.  Infected epithelial cells secrete chemokines \cite{Chan2005}, and contact with CD8 T cells also triggers secretion \cite{Zhao2000}. Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.  Previous work showed that efficient recruitment of T cells to the FOI is crucial for the eventual clearance of the virus \cite{Cerwenka1999, Kim2011}.  We extend this work to examine how the unique properties of the different strains of influenza may affect this process.

Here, we present an agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.  Previous models have shed light on the details of the primary and secondary immune response, some by applying ordinary differential equation (ODE) models to empirical data \cite{Miao2010, Saenz2010, Handel2008, Lee2009, Murillo2013}, and some by using spatial models to examine the interactions between dendritic cells and T cells in the lymph node \cite{Beltman2007, Zheng2008, Vroomans2012, Textor2014}.  We parameterize our model with \textit{in vitro} generated measurements of viral expansion dynamics as well as published data of host immune responses to flu infection.  The model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1, avian H5N1, and pandemic H1N1.  These three strains were selected for their differences in \textit{in vitro} replication rates \cite{Mitchell2011} as well as differences in \textit{in vivo} severity of human and animal pneumonia and mortality.  The model describes the dynamic, spatially explicit evolution of viral infections in the context of the developing CD8 T cell response.  Sensitivity analysis identifies parameters that are candidates for explaining the dramatic observed differences in pathogenicity among these different flu strains.

% The model design allows us to compare between influenza strains how the T cell response is affected by the viral and immune system characteristics.  The results show that efficient T cell search is crucial to viral clearance in the model and that the search process is dependent on more than just the replication rate of the virus, including the spatial landscape of the chemokine gradient and the location and density of virus-secreting cells.

The spatial nature of the model identifies physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.  Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.  The model suggests that the chemokine response of the host cells may actually hinder T cells' ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain.  Finally, the sensitivity analysis shows the isolated effects of each parameter in the model.  Based on this analysis, each parameter was classified as sensitive or not.  We find that most parameters can be individually varied across a broad range of values without strong effect on predicted outcomes.  Further, the parameter classification is consistent across each broad parameter type. Specifically, chemokine-related parameters are stable within biologically plausible ranges.  In contrast, parameters related to viral kinetics have a strong effect on model results.  We conclude that influenza infection dynamics are dominated by the replication properties of the virus and spatial properties of the infection, confirming that the higher replication rate of 2009 pH1N1 influenza obscured contributions of the chemotactic signals, leading to uncontrolled infection \textit{in silico}.

% fred used 'obscured' instead of overwhelmed

% The model predicts the control and clearance of the aH5N1 and sH1N1 strains and predicts a runaway infection for the pH1N1 strain.  These results are consistent with the previously reported replication rates of the three strains of influenza.

%  Infected cells secrete chemokine molecules which then diffuse across the infected region.  This process creates isolated regions, or local maxima, of high chemokine concentrations.

% While the local maxima do move and flatten, their speed is limited by their diffusion and decay rates.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}

% We are interested in how the interaction between the virus and the host's secondary immune response affects the virulence of the infection.  Because these interactions are difficult and expensive to observe \textit{in vivo}, we create a model \textit{in silico} to represent the system in question.  Because we are interested in individual interactions between the virus and the immune response, as well as the spatial nature of the spreading infection and the subsequent T cell search, we eschew traditional differential equation models \cite{Beauchemin2006a, Bauer2009} in favor of a spatial agent-based model (ABM).

@@

The model focuses on activated CD8 T cells migrating through the vascular and lymph networks, including their movement over tissue after extravasation (Fig.~\ref{fig:systemchart}).  The lung is represented as a two-dimensional sheet of healthy epithelial cells (S1.3).   Because we assume activated T cells descend at random into branches of the vascular network, T cells in the model are introduced uniformly at random across the modeled lung surface.  If cytokine signal is detected on the local endothelium the T cell remains in tissue and follows the chemotactic gradient to the FOI.  T cells that do not encounter cytokine recirculate to the lymph node where they reenter the bronchial vascular network.  Once inside the endothelium, T cells follow the chemokine gradient to localized areas of maximum chemokine concentration.  When a T cell contacts an infected epithelial cell it induces apoptosis (S1.4).  Otherwise, infected cells continue to secrete virus for a fixed time and then die.

The model is initialized with a single infected cell.  After the incubation period, the infected cell begins secreting virus and chemokine at rates determined by the virus strain and chemokine type (S2.1 and Table \ref{tab:strains}).  Virus from expressing cells diffuses locally, infecting neighboring cells. Chemokine diffuses from secreting cells, creating a spreading region of stimulation around the FOI. After five simulated days, representing lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a constant rate and travel through the vasculature to the tissue as described above.  These cell and molecular interactions and state transitions are depicted in Figure~\ref{fig:systemchart}.

@@

The model is implemented using the CyCells software \cite{Warrender2006} (\texttt{https://github.com/drewlevin/cycells}), a modeling platform for two- or three-dimensional agent-based simulations of the immune response (S1.1).  Due to computational constrains our model is scaled to the size of a mouse lung (approximately 100$cm^2$), a well-established practice in spatially explicit influenza models \cite{Miller2003, Allan2006, Ingulli2009}.  Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).

@@

In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead} \cite{bachem1996simulated, Beauchemin2005, Mitchell2011}. \texttt{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay. \texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die. \texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells. \texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour. \texttt{Dead} cells take up space and do not regenerate over the course of an infection.  Cell regrowth is not implemented as the rate of regrowth is unknown.

@@

Model parameters are listed in Tables \ref{tab:strains} and \ref{tab:parameters}. These values were taken from the literature when available.  Other parameters were determined experimentally, and some parameters were estimated using biologically plausible ranges.  All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior.

Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.  We extended a previously published differential equation model \cite{Mitchell2011} to obtain values for per-cell chemokine production rates (S2.1).  T cell production rates were derived from measured replication rates \textit{in vitro} \cite{Miao2010} using another differential equation model (S2.2). T cell production is assumed to be constant after day 5 \cite{MartIn-Fontecha2003}.

@@

In an effort to obtain per-cell chemokine and cytokine production rates, human epithelial cells were infected with influenza \textit{in vitro}.  Epithelial cell culture and supernatant collection was performed as described in \cite{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \cite{Mitchell2011}.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 $\mu$L aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.  Only IP-10, RANTES, and TNF$\alpha$ showed increases in production (Table S1).  TNF activity is not incorporated in the model.  Data for other chemokines and cytokines is not shown.

@@

Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).  For the avian and seasonal strains, all model runs cleared the virus by day 10.  Conversely, each simulation of the 2009 pandemic influenza led to runaway infection.

>>>>

\documentclass[preprint,10pt,authoryear]{elsarticle}

@@

% cite package, to clean up citations in the main text. Do not remove.
%\usepackage{cite}
\usepackage{natbib}

@@

\bibliographystyle{elsarticle-num-names}

@@

\usepackage{color}
\usepackage[usenames,dvipsnames,table]{xcolor}
%\usepackage{ulem}

@@

\journal{Theoretical Biology}

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{A spatial model of the efficiency of T cell search in the influenza-infected lung}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$,
Candice Clay$^{2}$, 
Judy Cannon$^{3}$,
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
\bf{3} Department of Molecular Genetics \& Microbiology Department of Pathology, University of New Mexico Health Sciences Center, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

%Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.  Viral clearance depends in part on the ability of T cells to locate the infection in time to prevent uncontrolled spread of the virus.  We use an agent-based spatial model to quantify how efficiently T cells locate and control different strains of influenza in the lung.  We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature.  The spatial model reveals challenges for T cell recruitment not apparent in standard differential equation models.  The model shows that virus kinetics largely determine the course of the infection, especially in the case of the pandemic flu strain, while T cell control of the seasonal flu strain is more important. Our results shed light on the key factors differentiating immune response against seasonal, avian, and pandemic flu.

%while T cell chemokine parameters have less effect.

@@

% ---- 300-ish word abstract

Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.  Subtle differences in how a given strain interacts with the immune system are likely a key factor in determining virulence.  One aspect of the interaction is the ability of T cells to locate the foci of the infection in time to prevent uncontrolled expansion.  Here, we develop an agent based spatial model to focus on T cell migration from lymph nodes through the vascular system to sites of infection. We use our model to investigate how different strains of influenza modulate this process.

We calibrate the model using viral and chemokine secretion rates we measure \textit{in vitro} together with values taken from literature.  The spatial nature of the model reveals unique challenges for T cell recruitment that are not apparent in standard differential equation models.  % -- Fred --
%We find that as plaques expand in size, the ratio of infected cells to dead cells decreases, making it difficult for T cells to encounter infected cells efficiently.  Further, differences in the diffusion rates of chemokines and virions can misdirect T cells to areas that no longer contain infected epithelial cells.  While our sensitivity analysis identifies viral parameters as the most significant factor in determining the course of the infection, we are also able to determine an upper bound of T cell search time using the model.  Finally, we use the model parameters to determine an upper bound on T cell search time in the model.  Our results shed light on the key factors differentiating immune response against seasonal, avian, and pandemic flu.
% -- Fred , after p2s2 -- 
In this model comparing three influenza viruses, plaque expansion is governed primarily by the replication rate of the virus strain, and the efficiency of the T cell search-and-kill is limited by the density of infected epithelial cells in each plaque.  Thus for each virus there is a different threshold of T cell search time after which recruited T cells are unable to control further expansion.  Future models could use this relationship to more accurately predict control of the infection.

% ---- end 300-ish word abstract

@@

@@

Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \citep{Who2009}.  Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the the mechanisms that control this variability are not well understood.  Pathogenicity is a function of the virus and its interaction with the host immune response.

Identifying the critical components of this complex process, and how they interact, is key to understanding viral pathogenicity and designing therapeutic interventions.  We address this challenge through computational modeling, which allows us to study the relative contributions of different aspects of the host immune response and to account for strain-specific viral dynamics in the model.  Several earlier theoretical models have been proposed to study specific aspects of influenza infections. Ordinary differential equation (ODE) models that are fit to empirical data \citep{Handel2008, Lee2009, Miao2010a, Saenz2010, Murillo2013, Crauste2015, Price2015} have elucidated viral population level dynamics but ignore localized spatial effects.  Spatial models, on the other hand, have been developed to examine interactions between dendritic cells and T cells in the lymph node \citep{Beauchemin2005, Beltman2007, Zheng2008, Mirsky2011, Celli2012, Vroomans2012, Textor2014}, but have not been extended to consider local conditions in the lung.   In this paper, we examine how the unique properties of three different strains of influenza, in the presence of chemokines, affect T-cell search in the human lung.

Specifically, we focus on the interactions between activated antigen-specific CD8 T cells, cytokines, and replicating influenza virus.  Chemotactic proteins are known to enhance T cell recruitment in both acute infections and chronic inflammatory diseases \citep{Gunn1998, Medoff2005, Okada2005, Castellino2006, Bromley2008}. Infected epithelial cells secrete chemokines, especially upon contact with CD8 T cells  \citep{Zhao2000, Chan2005}. Previous work showed that efficient recruitment of T cells to the Focus Of Infection (FOI) is crucial for the eventual clearance of the virus\citep{Cerwenka1999, Kim2011}.   As measuring real-time T cell movement in \textit{in vivo} lung tissue is not currently viable, study these localized processes is aided by spatial modeling of how T cells interact with the FOI and the chemotactic environment.  Therefore, instead of developing a comprehensive immune system model, we present a spatially explicit agent-based model (ABM) to describe T cell interactions with chemotactic signals and a dynamically growing plaque.  Using the model, we investigate the pathogenic potential of the three  influenza strains: seasonal H1N1 (sH1N1), Avian H5N1 (aH5N1), and pandemic H1N1 (pH1N1). These three strains were selected for their differences in \textit{in vitro} replication rates \citep{Mitchell2011} as well as differences in \textit{in vivo} severity of human and animal pneumonia and mortality.

The model is parameterized with values taken from the literature when available.  Because chemokine secretion rates are central to our model and appropriate data values are not available, we estimate the parameters using data from physical experiments and an earlier ODE model \citep{Mitchell2011}.   To obtain the physical data, wells of human epithelial cells were infected with one of the three strains and then measured for virus and cytokine concentrations over a 48-hour period.   Of the cytokines tested, only the chemokines CXCL10 (IP-10) and CCL5 (RANTES) were expressed differently across the three influenza strains.   IP-10 and RANTES are both chemokines that normally attract T cells to the FOI \citep{Hoji2005, Groom2011a}.  These results suggest that the two chemokines may contribute to the immune system's varying ability to control different strains of influenza, and were thus included in our model.

The spatial model and the supporting experiments reveal several interesting and underappreciated phenomena.  The model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection. Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration. The model suggests that the chemokine response of the host cells can hinder T cells’ ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain. A sensitivity analysis of the model parameters isolated the effects of each parameter in the model.  The analysis shows that most model parameters can vary across a broad range of values without strong effect on predicted outcomes.  Further analysis shows model predictions remain unchanged within broad categories of parameter type (e.g., all parameters related to chemokines, parameters related to viral kinetics, etc.).  For example, the chemokine-related parameters are stable in that they do not affect the model behavior within biologically plausible ranges. In contrast, parameters related to viral kinetics significantly alter model predictions.  Finally we examine the `window of control' which establishes an upper bound on the time it takes T cells to arrive at the FOI while still clearing the infection.  We find that this value is consistent with our sensitivity analysis and helps explain the challenge of controlling the highly virulent pH1N1 strain.

%We conclude that influenza infection dynamics are dominated by the replication properties of the virus and spatial properties of the infection. The higher replication rate of 2009 pH1N1 influenza dominated the effect of chemotactic signals, leading to uncontrolled infection \textit{in silico}.

\section*{Methods and Models}

% We are interested in how the interaction between the virus and the host's secondary immune response affects the virulence of the infection.  Because these interactions are difficult and expensive to observe \textit{in vivo}, we create a model \textit{in silico} to represent the system in question.  Because we are interested in individual interactions between the virus and the immune response, as well as the spatial nature of the spreading infection and the subsequent T cell search, we eschew traditional differential equation models \citep{Beauchemin2006a, Bauer2009} in favor of a spatial agent-based model (ABM).

@@

The model focuses on activated CD8 T cells migrating through the vascular and lymph networks, including their movement over tissue after extravasation (Fig.~\ref{fig:systemchart}).  While the volume of the lung consists of spherical alveoli, the spread of influenza throughout the lung occurs through the interconnected bronchiole-alveoli complex, which can be viewed as a connected monolayer.  Thus, we represent the lung as a two-dimensional sheet of healthy epithelial cells (S1.3).   Because we assume activated T cells descend at random into branches of the vascular network, T cells in the model are introduced uniformly at random across the modeled lung surface.  If cytokine signal is detected on the local endothelium the T cell remains in tissue and follows the chemotactic gradient to the FOI.  T cells that do not encounter cytokine recirculate to the lymph node where they reenter the bronchial vascular network.  Once inside the endothelium, T cells follow the chemokine gradient to localized areas of maximum chemokine concentration.  When a T cell contacts an infected epithelial cell it induces apoptosis (S1.4).  Otherwise, infected cells continue to secrete virus for a fixed time and then die.

The model is initialized with a single infected cell.  After the incubation period, the infected cell begins secreting virus and chemokine at rates determined by the virus strain and chemokine type (Table \ref{tab:strains}).  Virus from expressing cells diffuses locally, infecting neighboring cells. Chemokine diffuses from secreting cells, creating a spreading region of stimulation around the FOI. After five simulated days, representing lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a constant rate and travel through the vasculature to the tissue as described above.  These cell and molecular interactions and state transitions are depicted in Figure~\ref{fig:systemchart}.

@@

The model is implemented using an updated version of the CyCells software \citep{Warrender2006},
(\texttt{github.com/drewlevin/cycells}),
a modeling platform for two- or three-dimensional agent-based simulations of the immune response (S1.1).  Due to computational constrains our model is scaled to the size of a mouse lung (approximately 100$cm^2$), a well-established practice in spatially explicit influenza models \citep{Miller2003, Allan2006, Ingulli2009}.  Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).

@@

In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead} \citep{bachem1996simulated, Beauchemin2005, Mitchell2011}. \texttt{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay. \texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die. \texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells. \texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour. \texttt{Dead} cells take up space and do not regenerate over the course of an infection.  Cell regrowth is not implemented as the rate of regrowth is unknown.

@@

Model parameters are listed in Tables \ref{tab:strains} and \ref{tab:parameters}. These values were taken from the literature when available.  Other parameters were determined by matching empirical data to ODE models, and some parameters were estimated using biologically plausible ranges.  All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior (Table~\ref{tab:sensitivity}).

Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.  We extended a previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.  T cell production rates were derived from measured replication rates \textit{in vitro} \citep{Miao2010a} using another differential equation model. T cell production is assumed to be constant after day 5 \citep{MartIn-Fontecha2003}.

\subsection*{Models for Parameter Estimation}

We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \citep{Mitchell2011} Eq. 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.  Initial population sizes and parameter values are taken from the previous study.
{\footnotesize
\begin{equation*}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_{1 \tau_2} \\
\dot{C} &= r I_{1 \tau_3} - d C \\
\end{aligned}
\tag{Eq. 1}
\label{eq:dde}
\end{equation*}
\vspace{.05in}
}

$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.  Table S2 summarizes population and parameter values and descriptions.  Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1.  The results of these fits are discussed in the main paper and shown in Table 1.  IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).

%\subsection*{Chemokine Secretion}

%To provide estimates of chemokine concentrations and secretion rates present in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.~\ref{fig:data}).  The dynamic viral loads at these intervals in these cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus have been reported previously \citep{Mitchell2011}.  IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..

%To estimate per-cell chemokine production rates, we extended the delay differential equation (DDE) model of Ref. \citep{Mitchell2011} to include chemokine production from infected cells (Eq.~S1).  Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).  The resulting chemokine production values were used in the CyCells ABM by combining IP-10 and RANTES into one aggregate rate but also reported separately (Fig.~S2).  Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.  There was no positive correlation between viral production and induced chemokine production across the three strains.

% \subsection{Estimating the T cell production rate}

We calculate the rate of production, $\sigma$, of CD8 T cells using a differential equation model from \citep{Miao2010a}.  The equations model T cell production and subsequent search over an area of infected lung tissue.

{\footnotesize
\begin{equation*}
\label{eq:miao}
\begin{aligned}
\dot{N_{c}} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} & \pi r'^{2} &= a \sqrt{I} + b \\
\dot{N'_{f}} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
\dot{N_{f}} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
\dot{T} &= \rho T -\beta TV \\
\dot{I} &= \beta TV - \delta I - k_{e} N_{f} I \\
\dot{V} &= pI - \beta TV - \gamma (t) V \\
\gamma (t) &= \left\{ \begin{array}{rcl}
	1/\mbox{day} & \mbox{,}  & t < 5  \\
	3/\mbox{day} & \mbox{,} & t \geq 5  
	\end{array}\right. \\
\end{aligned}
\tag{Eq. 2}
\end{equation*}
}
\vspace{0.5in}

$N_{c}$ is the number of circulating activated antigen-specific CD8 T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines, $N_{f}$ is the number of circulating T cells that have found an infected region. $T$ is the number of uninfected target cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum. The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$). The lung is modeled as a circular region with radius $R$. The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells. See Fig. 1 for more details about the model.  The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.  $a$ and $b$ were fit using 12 experimental runs of a spatial model implemented in CyCells.

Circulating CD8 T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$ and circulate to the $N'_{f}$ population over a time defined by $t_{rc}$. We assume these circulating cells transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.

The average time an exiting CD8 T cell takes to migrate to the infected region is equal to the difference in the radii between the two regions ($r' - r$) times the T cell speed, $v_{tcell}$.  Circulating cells that are in the chemokine expressing region ($N'_{f}$) move into the infected region after performing chemotaxis for that time. Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection. Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T cells ($N_{f}$) at a rate $k_{e}$. Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells. Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection). The initial viral titer was initialized to 10,000 PFUs and the initial number of target cells was one million. The initial number of infected cells is assumed to be zero. Parameter values are listed in Table S3.  This ODE was fit to data taken from \citep{Miao2010a} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.  The final value was found to be 1,257 per hour.

@@

In an effort to obtain per-cell chemokine and cytokine production rates, human epithelial cells were infected with influenza \textit{in vitro}.  Epithelial cell culture and supernatant collection was performed as described in \citep{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \citep{Mitchell2011}.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 $\mu$L aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.  Only IP-10, RANTES, and TNF$\alpha$ showed increases in production (Table S1).  TNF activity is not incorporated in the model.  Data for other chemokines and cytokines is not shown.

@@

Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three influenza strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).  For the avian and seasonal strains, all model runs cleared the virus by day 10.  Conversely, each simulation of the 2009 pandemic influenza led to uncontrolled infection.
